\section[Path-Independent Vector Fields]{Path-Independent Vector Fields and the Fundamental Theorem of
Calculus for Line Integrals} \label{S:12.4.FTCLI}


\vspace*{-14 pt}
\framebox{\hspace*{3 pt}
\parbox{6.25 in}{\begin{goals}
\item How can we characterize the vector fields $\vF$ for which
  $\int_C\vF\cdot d\vr$ has the same value for every oriented path
  from a point $P$ to a point $Q$?
\item What special properties do gradient vector fields have?
\end{goals}} \hspace*{3 pt}}

\subsection*{Introduction}

In some of the activities and examples in this chapter we have
encountered situations where $C_1$ and $C_2$ are different oriented curves from a
point $P$ to a point $Q$ and $\int_{C_1}\vF\cdot d\vr =
\int_{C_2}\vF\cdot d\vr$. In this section, we explore vector fields
which have the property that for all points $P$ and $Q$, if $C_1$ and
$C_2$ are oriented paths from $P$ to $Q$, then $\int_{C_1}\vF\cdot
d\vr = \int_{C_2}\vF\cdot d\vr$.

\input{previews/12.4.PA1}

\subsection*{Path-Independent Vector Fields}

We say that a vector field $\vF$ defined on a domain $D$ is
\textbf{path-independent} \index{pathindependent!definition} if
$\int_{C_1}\vF\cdot d\vr = \int_{C_2}\vF\cdot d\vr$ whenever $C_1$ and
$C_2$ are oriented paths in $D$ having the same initial point and same
terminal point. 

We've already encountered some situations where we had evidence that a
vector field was path-independent, but since the definition requires
that for every pair of points in the domain and every possible pair of
paths from oine point to the other we must get the same value, it
doesn't appear that verifying a vector field is path-independent is an
easy task. Fortunately, one familiar class of vector fields can be
shown to be path-independent. Let $f\colon \R^3\to \R$ be a function
for which $\nabla f$ is continuous on a domain $D$. Suppose that $P$
and $Q$ are points in $D$ and let $C$ be a smooth oriented path from $P$ to
$Q$. Let's take a look at $\int_C\nabla f \cdot d\vr$ by fixing an arbitrary
parameterization $\vr(t)$ of $C$, $a\leq t \leq b$. Since $\nabla
f(\vr(t)) = \langle f_x(\vr(t)),f_y(\vr(t)),f_z(\vr(t))\rangle$, we know that
\[\int_C\nabla f\cdot d\vr  = \int_a^b \nabla f(\vr(t))\cdot
\vr'(t)\, dt = \int_a^b \langle f_x(\vr(t)), f_y(\vr(t)),
f_z(\vr(t))\rangle \cdot \vr'(t)\, dt.\]
If $\vr(t) = \langle x(t), y(t), z(t)\rangle$, then the integrand
above is
\begin{multline*}\langle f_x(\vr(t)), f_y(\vr(t)), f_z(\vr(t))\rangle\cdot \langle
x'(t),y'(t),z'(t)\rangle =\\ f_x(x(t),y(t),z(t))x'(t) +
f_y(x(t),y(t),z(t))y'(t) + f_z(x(t),y(t),z(t))z'(t).\end{multline*}
Notice that this is exactly what the chain rule tells us
$\ds\frac{d}{dt} f(\vr(t))$ is equal to. Therefore,
\[\int_C\nabla f\cdot d\vr = \int_a^b \frac{d}{dt} f(\vr(t))\, dt =
f(\vr(b)) - f(\vr(a)) = f(Q) - f(P).\]
In other words, gradient vector fields are path-independent vector
fields, and we can evaluate line integrals of gradient vector fields
by using a potential function. (Technically the argument above assumed
that $C$ was smooth, but we can replace $C$ by a piecewise smooth
curve by splitting the line integral up into the sum of finitely many
line integrals along smooth curves.)

This result is so important that it is frequently called the
Fundamental Theorem of Calculus for Line integrals, because of its
similarity to the Fundamental Theorem of Calculus, which can be
written as
\[\int_a^b f'(x)\, dx = f(b) - f(a).\]
\vspace*{5pt}
\nin \framebox{\hspace*{3 pt}
  \parbox{6.25 in}{\textbf{Fundamental Theorem of Calculus for Line
      Integrals} Let $f$ be a function for which $\nabla f$ is
    continuous on a domain $D$. If $P$ and $Q$ are points in $D$ and
    $C$ is a piece-wise smooth oriented path from $P$ to $Q$ in $D$, then
  \[\int_C \nabla f\cdot d\vr = f(Q) - f(P).\]} \hspace*{3 pt}} \vspace*{5pt}

\input{activities/12.4.Act1}

By following the methodology laid out in Preview
Activity~\ref{PA:12.4} to show that a given vector field is a gradient
vector field, the Fundamental Theorem of Calculus for Line Integrals
makes evaluating a large number of line integrals simpler now.

\input{activities/12.4.Act1}

\input{activities/12.4.Act2}
%\clearpage %% Temporary just to make a handout
\subsection*{Line Integrals Along Closed Curves}

Recall that we call an oriented curve $C$ \textbf{closed} if it has
the same initial and terminal point. A typical example of a closed
curve would be a circle (with orientation), but we could also consider
something like the square with vertices $(1,1)$, $(-1,1)$, $(-1,-1)$,
and $(1,-1)$, oriented clockwise (or counterclockwise).

\input{activities/12.4.Act3}

From Activity~\ref{A:12.4.3}, we now know that $\vF$ is path-independent if and only if $\oint_C\vF\cdot d\vr = 0$ for all closed
curves $C$ in the domain of $\vF$. Although this is not a terribly
useful way to show that a vector field is path-independent, it can be
a useful way to show that a vector field is \emph{not}
path-independent: find a closed curve around which the circulation is
not zero.

\input{activities/12.4.Act4}

\subsection*{What other vector fields are path-independent?}

Recall that in single variable calculus, the Second Fundamental
Theorem of Calculus tells us that given a constant $c$ and a
continuous function $f$, there is a unique function $A(x)$ for which
$A(c) = 0$ and $A'(x) = f(x)$. In particular, $A(x) = \int_c^x f(t)\,
dt$ is this function. We are about to investigate an analog of this result for
path-independent vector fields, but first we require two additional
definitions.

If $D$ is a subset of $\R^2$ or $\R^3$, we say that $D$ is
\textbf{open}\index{open!definition} provided that for every point in $D$, there
is a disc (in $\R^2$) or ball (in $\R^3$) centered at that point such
that every point of the disc/ball is contained in $D$. For example,
the set of points $(x,y)$ in $\R^2$ for which $x^2+y^2 < 1$ is open,
since we can always surround any point in this set by a tiny disc
contained in the set. However, if we change the inequality to
$x^2+y^2\leq 1$, then the set is not open, as any point on the circle
$x^2+y^2=1$ cannot be surrounded by a disc contained in the set; any
disc surrounding a point on that circle will contain points outside
the set, that is with $x^2+y^2>1$. We will also say that a region $D$
is \textbf{connected}\index{connected!definition} provided that for
every pair of points in $D$, there is a path from one to the other
contained in $D$.

\input{activities/12.4.Act5}

We summarize the result of Activity~\ref{A:12.4.5} below.

\nin \framebox{\hspace*{3 pt}
  \parbox{6.25 in}{\textbf{Path-Independent Vector Fields} If $\vF$ is
  a path-independent vector field on an open, connected domain $D$,
  then $\vF$ is conservative (or a gradient vector field) on
  $D$. Furthermore, if $P$ is a point in $D$ and $f(P)$ is fixed, then
  the function \[f(Q) = f(P) + \int_C\vF\cdot d\vr\]
  (where $Q$ is a point in $D$ and $C$ is an oriented curve from $P$ to
  $Q$ in $D$) is a 
  potential function for $\vF$.} \hspace*{3 pt}} \vspace*{5pt}


%\clearpage %% Temporary just to make a handout

%\nin \framebox{\hspace*{3 pt}
%\parbox{6.25 in}{
\begin{summary}
\item Gradient vector fields are path-independent, and if $C$ is an
  oriented curve from $(x_1,y_1)$ to $(x_2,y_2)$, then $\int_C\nabla f\cdot
  d\vr = f(x_2,y_2)-f(x_1,y_1)$, with the analogous result holding if
  $f$ is a function of three variables.
\item A vector field is path-independent if and only if the
  ciruclation around every closed curve in its domain is $0$.
\item If a vector field $\vF$ is path-independent, then there exists a
  function $f$ such that $\nabla f = \vF$. That is, $\vF$ is a
  conservative or gradient vector field.
\end{summary}
%} \hspace*{3 pt}}

\nin \hrulefill

%\input{exercises/9.2.Vectors(Ex)}

\clearpage

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "0_AC_MV"
%%% End:
